---
title: 'Data Cleaning & Processing'
author: "Team 17"
date: 'Due 2024-11-07'
output:
  pdf_document: default
  html_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Data Cleaning and Processing

This R Markdown document details the steps taken to ingest, clean, and integrate the three macroeconomic time-series datasets: Federal Funds Rate (FEDFUNDS), Unemployment Rate (UNRATE), and GDP Growth Rate.

The primary cleaning goals are to ensure proper data types (especially for the date column), assign descriptive column names, and merge the datasets for unified analysis.

# Process Federal Funds Rate

```{r}
fedfunds_processed <- clean_data(
  file_path = "FEDFUNDS.csv", 
  new_col_name = "Fed_Funds_Rate", 
  original_col_name = "FEDFUNDS"
)
```

# Process Unemployment Rate

```{r}
unrate_processed <- clean_data(
  file_path = "UNRATE.csv", 
  new_col_name = "Unemployment_Rate", 
  original_col_name = "UNRATE"
)
```

# Process GDP Growth Rate
```{r}
gdp_processed <- clean_data(
  file_path = "GDP_GROWTH_RATE.csv", 
  new_col_name = "GDP_Growth_Rate", 
  original_col_name = "A191RL1Q225SBEA"
)
```

# Federal Funds Rate Data Inspection
```{r}
head(fedfunds_processed)
```

# Federal Unemployment Rate Data Inspection
```{r}
head(unrate_processed)
```

# Growth Rate Data Inspection
```{r}
head(gdp_processed)
```

# Merging the 3 datasets
```{r}
combined_data <- merge_macro_data(
    fedfunds_processed, 
    gdp_processed, 
    unrate_processed
)
```

# Combined Data Inspection
```{r}
head(combined_data)
```
