---
title: 'Analyzing the Relationship Between Interest Rates and Economic Growth Across Varying Financial Crises'
author: "Team 17: Gokul Giridharan, Connor Huang, Arjun Singh"
link: "https://github.com/snazzy77/STAT107_Team-17_Financial-Crisis.git"
date: 'Due 2025-12-05'
output:
  pdf_document: default
  html_document: default
---

**GitHub Repository:** <https://github.com/snazzy77/STAT107_Team-17_Financial-Crisis>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width = 4, fig.height = 3)
source("00_requirements.R")
source("01_funct_DataCleaning.R")
source("02_funct_plots.R")
load("final_analysis_environment.RData")
```

# Abstract

We will be analyzing the relationship between interest rates and economic growth/indicators across financial crises that occurred in 2001, 2008, and 2020 (the dot-com crisis, the great recession and the COVID-induced recession). We will specifically analyze GDP growth and unemployment during the three major U.S. financial crises and we will determine whether changes in interest rates reveal relationships with GDP growth and unemployment across economic disasters. According to our preliminary analysis, interest rates and GDP growth seem to move in the same direction; however, the strength and significance of this relationship varies across crises. In this project, we analyze the relationship between U.S. interest rates and unemployment during both crisis and non-crisis periods. We aim to determine whether monetary policy reacts more strongly to labor market conditions during recessions.

# Introduction

The ultimate goal of this project is to explore how interest rates interact with key economic indicators such as GDP growth and unemployment during varying financial crises. Understanding this relationship will lead to a better comprehension of how monetary policy affects the country’s economy, which will benefit economists or politicians who are interested in how the Federal Reserve’s actions affect the economy. We will fit separate linear regression models for each recession to test whether interest rates are significant predictors of GDP growth, and then evaluate whether a unified model across all periods can describe the data effectively and accurately.

# Data

We filtered the dataset to include only complete observations, removed missing values, and ensured both variables were aligned by month and time period. Data was split into crisis vs non-crisis subsets for modeling. Our dataset will be a combination of public economic data from the Federal Reserve Economic Data (FRED) database. Key variables we will utilize include:

-   Date: Quarter (YYYY-Q)

-   Interest_Rate: Federal Funds Effective Rate (%)

-   GDP_Growth: Quarterly GDP growth rate (%)

-   Unemployment_Rate: Civilian Unemployment Rate (%)

To clean the data we ran a function to convert the observation date to the Date data type using as.Date(). We did this for all the data sets. At the same time we also changed the column names to something more appropriate. Lastly we wanted to be able to make better plots with the data so we decided to merge them.

# Visualization

```{r, echo = FALSE}
summary(combined_data)
```

This summary table provides descriptive statistics for all of the variables in the combined dataset, including Federal Funds Rate, GDP Growth Rate, and Unemployment Rate. It shows that GDP growth mostly falls between 0% and 5%, while federal funds rates are generally below 10%.

```{r, echo = FALSE}
plot_histogram(
    data = combined_data,
    column_name = "GDP_Growth_Rate",
    title_suffix = "GDP Growth Rate"
)

```

This visualization is a histogram of the GDP growth rate. We can see that most U.S. GDP growth rates fall between 0% and 5%, indicating that the economy typically expands at a modest pace. Occasional negative values represent recession periods where economic output declined. A few extreme highs and lows appear, reflecting rare events such as severe downturns or unusually strong recoveries.

```{r, echo = FALSE}
plot_density(
    data = combined_data,
    column_name = "Fed_Funds_Rate",
    title_suffix = "Federal Funds Rate",
    color = "purple"
)

```

This graph shows a Density plot of federal funds rate. We can see that U.S. federal funds rates most frequently fall in the low-to-moderate range, with the highest concentration around roughly 4–6%. Rates above 10% occur much less often, indicating that very high interest-rate environments are historically rare. Overall, the distribution is right-skewed, meaning low and moderate rates are far more common than extremely high ones.

```{r, echo = FALSE}
plot_scatter(
    data = combined_data,
    x_col = "Unemployment_Rate",
    y_col = "Fed_Funds_Rate",
    main_title = "Federal Funds Rate vs. Unemployment Rate",
    x_label = "Unemployment Rate (%)",
    y_label = "Fed Funds Rate (%)"
)

```

This graph is a scatter plot showing the Federal Funds Rate vs. Unemployment Rate. It shows the relationship between the U.S. federal funds rate and the unemployment rate. There is no strong linear relationship, though higher unemployment rates often coincide with lower federal funds rates, reflecting monetary policy easing during weak labor markets. Overall, most observations cluster around moderate unemployment (4–8%) with fed funds rates below 10%.

```{r, echo = FALSE}
plot_scatter(
    data = combined_data,
    x_col = "Unemployment_Rate",
    y_col = "GDP_Growth_Rate",
    main_title = "GDP Growth Rate vs. Unemployment Rate",
    x_label = "Unemployment Rate (%)",
    y_label = "GDP Growth Rate (%)"
)
```

This visualization is a scatterplot of the GDP Growth Rate vs. Unemployment Rate. It shows the relationship between U.S. GDP growth and unemployment rates. There is no clear linear pattern, though higher unemployment generally corresponds to lower or negative GDP growth, consistent with economic downturns. Most observations cluster around moderate unemployment (4–8%) and modest GDP growth (0–5%), with only a few extreme outliers.

```{r, echo = FALSE}
plot_scatter(
    data = combined_data,
    x_col = "Fed_Funds_Rate", # Using Fed Funds Rate as the predictor (x)
    y_col = "GDP_Growth_Rate",
    main_title = "GDP Growth Rate vs. Federal Funds Rate",
    x_label = "Fed Funds Rate (%)",
    y_label = "GDP Growth Rate (%)"
)
```

This scatter plot visualizes GDP growth vs. federal funds rate. There is a slight positive trend during non-recession periods, while low rates during recessions correspond to negative GDP growth.

```{r, echo = FALSE}
plot_scatter2(
    data = combined_data,
    x_col = "Unemployment_Rate",
    y_col = "Fed_Funds_Rate",
    main_title = "Unemployment vs. Fed Funds Rate (Colored by Crisis)",
    x_label = "Unemployment Rate (%)",
    y_label = "Fed Funds Rate (%)",
    color_col = "Is_Crisis" # Colors points based on the new variable
)
```

This scatter plot shows unemployment vs. federal funds rate, with points colored by crisis periods. During recessions, higher unemployment aligns with lower rates, indicating the Fed’s economic response.

```{r, echo = FALSE}
plot_scatter(
    data = combined_data,
    x_col = "Fed_Funds_Rate",
    y_col = "GDP_Growth_Rate",
    main_title = "GDP Growth Rate vs. Federal Funds Rate",
    x_label = "Fed Funds Rate (%)",
    y_label = "GDP Growth Rate (%)"
)
```

# Linear Model

```{r, echo = FALSE}
# Output test results (these objects are saved for the report)
cor_unrate_fedfunds
cor_gdp_fedfunds
```

These correlation values indicate the linear relationships between unemployment and federal funds rate (cor_unrate_fedfunds) and GDP growth and federal funds rate (cor_gdp_fedfunds). Essentially, negative correlation for unemployment demonstrates rate cuts during high unemployment, while GDP correlation shows how rates track economic growth.

```{r, echo = FALSE}
# Print summaries for documentation/review
summary(lm_crisis)
summary(lm_non_crisis)
```

These linear model summaries show how well federal funds rate predicts GDP growth during crisis (lm_crisis) and non-crisis (lm_non_crisis) periods. Coefficients, standard errors, and p-values are all indicators of whether or not interest rates significantly predict GDP growth as well as if relationships differ by economic conditions.

# Analysis

Using U.S. data on interest rates, GDP growth, and unemployment, we began by visually exploring how these variables behaved during the three major recessions of 2001, 2008, and 2020. The density plot of the federal funds rate shows that most rates tend to fall between about 4–6%, while very high rates are rare. The histogram of GDP growth suggests that the U.S. usually experiences modest growth, but growth becomes negative during recession years. When comparing interest rates and unemployment on a scatterplot, we see that periods of higher unemployment typically align with lower interest rates, which is consistent with the Federal Reserve lowering rates to stimulate the economy. The scatterplot of GDP growth and unemployment shows a wide spread of points, meaning higher unemployment does not always correspond directly to lower GDP growth, though extreme unemployment does tend to occur during times of very poor growth.

To quantify these relationships, we calculated Pearson correlations across all periods: Unemployment Rate vs. Fed Funds Rate: correlation = 0.053, p-value = 0.376 This indicates almost no linear association across the data set.

```         
GDP Growth Rate vs. Fed Funds Rate: correlation = -0.081, p-value = 0.174
  While just slightly negative, the correlation is weak and not statistically significant.
```

We then ran separate linear regressions for crisis and non-crisis periods: The R2 value for crisis periods was 0.457, meaning that during recessions, higher unemployment reliably predicts lower interest rates. The R2 value for non-crisis periods was 0.022, meaning that unemployment explains essentially none of the variability in interest rates.

The results that we got from the regressions confirm what our visualizations suggested: during recessions, monetary policy shows a stronger and more predictable relationship with unemployment, while in non-crisis periods, the connection is weak and inconsistent.

Going forward, we plan to run separate simple linear regression models for each recession to test whether interest rates help predict GDP growth during those periods. This will allow us to see whether the strength of the relationship between the two variables changes across crises. If interest rates are significant predictors, it would support the idea that monetary policy plays an important role during economic downturns. We also plan to examine whether unemployment shows any statistical relationship with interest rates or GDP growth by looking at correlations and summary statistics.

# Conclusions

All in all, our research led us to make several conclusions in regards to interest rates, GDP growth, and unemployment: - During recessions, the Federal Reserve lowers interest rates in response to rising unemployment. This relationship is statistically significant and explains a moderate proportion of the variability in rates. - Across all periods, unemployment on its own does not accurately predict interest rates or GDP growth. Economic dynamics in stable periods are influenced by multiple factors aside from just unemployment. - Recessions, specifically the COVID-19 pandemic, introduce extreme values in GDP growth, unemployment, and interest rates. For this reason, it is important to model each economic decline individually rather than using a singular model.

In the future, we may combine the data from all three recessions into a single model and compare it to the individual recession models. If the combined model performs worse, this would confirm that each crisis behaves differently and needs to be studied individually. Overall, our findings suggest that interest rates move in a predictable pattern during economic downturns. However, these relationships are complex and vary depending on the crisis. Therefore, analyzing each recession individually provides a better understanding of the impact of monetary policy on the economy. These results demonstrate that the Federal Reserve prioritizes employment stabilization during recessions but balances many additional factors in stable periods such as inflation and financial market conditions. Each crisis carries a unique economic structure, so treating recessions as identical events can lead to misleading conclusions about monetary policy effectiveness. One promising extension would be to incorporate inflation, consumer spending, and financial market indicators to determine whether a richer multivariable model can improve predictive accuracy. Time-series forecasting methods could also be used to evaluate whether monetary policy shifts can be predicted ahead of economic downturns.

# Contributions

Gokul Giridharan: Did the Data Filtering and Merging. Worked on coding the Visualizations and Models.

Connor Huang: Found the Datasets. Worked on the Analysis and Conclusion.

Arjun Singh: Worked on the descriptions for the visualizations and the Analysis and Conclusions. Also helped with the coding for the visualizations.
